Graded Call-By-Push-Value

### 2020-07-10 popl21/popl21.tex
[ popl21 ] related work and conclusions

Coeffect type systems
\citep{brunel:esop14,orchard:icfp14,ghicaSmith:esop14} have been
developed to track resource consumption by attaching usage information
to each variable in the typing context.  Coeffect typing generalizes
linear typing \citep{girard:linear} to quantitative typing
\citep{sergeyVytiniotisPeytonJones:popl14,mcBride:wadler60,atkey:lics18}
and subsumes sensitivity analysis \citep{reedPierce:icfp10} and static
information control flow \citep{volpano:jcs96} aka security typing.
Coeffects can be modeled by a comonad $\D[r]$ graded over resource
qualifiers $r$ drawn from a preordered semiring.  For the sake of
coeffect-graded CBPV, we generalize graded comonads to graded comonad
coalgebras.  Value types $P$ and contexts $\Gamma$ are then
interpreted as such coalgebras.  Value typing
$\qValTy \gamma \Gamma v P$ and computation typing
$\qCompTy \gamma \Gamma t N$ are based on linear typing an come
equipped with a resource context~$\gamma$.  Semantically, values are
interpreted as morphisms $\Gamma_{r\gamma} \to P_r$ allowing the
``multiplication'' or \emph{scaling} of a value by $r$ % natural in $r$
and computations as morphisms $\Gamma_\gamma \to N$.

@@

\item We further introduce graded comonad coalgebras and a
  coeffect-graded version of CBPV in Section~\ref{sec:coeff}.  We give
  its denotational and operational semantics and adapt Atkey and
  Wood's substitution theorem \citeyearpar{atkeyWood:types19} to this
  version of CBPV.

@@

The matrix $\Psi : \R^{\dom \Gamma \times \dom \Delta}$ acts as a
linear map
$(\delta : \R^{\dom \Delta}) \mapsto (\delta\Psi : \R^{\dom \Gamma})$
and allows us to state the substitution theorem.
\begin{theorem}[Substitution preserves typing]
  Let $\SubstTy \Psi \Gamma \sigma \Delta$.
  \begin{enumerate}
  \item If $\qValTy \delta \Delta v P$ then $\pValTy {\delta\Psi} \Gamma {v\sigma} P$.
  \item If $\qCompTy \delta \Delta t N$ then
    $\pCompTy {\delta\Psi} \Gamma {t\sigma} N$.
  \end{enumerate}
\end{theorem}
\begin{proof}
  Straightforward adaptation of \citet{atkeyWood:types19}.
\end{proof}
\begin{corollary}[Single substitution preserves typing]
  \label{cor:tysinglesub}
  \bla
  \begin{enumerate}
  \item If $\qValTy \delta \Gamma v P$ and
      $\ValTy {\qext \gamma \Gamma r x P} w {P'}$
      then $\pValTy {\gamma + r\delta} \Gamma {\qsubst v r x w} {P'}$.
  \item If $\qValTy \delta \Gamma v P$ and
      $\oCompTy {\qext \gamma \Gamma r x P} t {N}$
      then $\pCompTy {\gamma + r\delta} \Gamma {\qsubst v r x t} {N}$.
  % \item If $\qCompTy \delta \Gamma u N$ and
  %     $\oCompTy {\qext \gamma \Gamma r y N} t {N'}$
  %     then $\pCompTy {\gamma + r\delta} \Gamma {\qsubst u r y t} {N'}$.
  \end{enumerate}
\end{corollary}
% %\begin{remark}
% Analogous statements are available for the substitution into values
% $\ValTy {\qext \gamma \Gamma r x P} w {P'}$ and
% $\ValTy {\qext \gamma \Gamma r y N} w {P}$, however, they are not
% invoked in the proof of subject reduction.
% %\end{remark}
\begin{corollary}[Subject reduction]
  $\beta$-reduction of coeffect-graded CBPV preserves types.
\end{corollary}
In the pure (effect-free) version, every computation at value type
returns a value.
\begin{conjecture}[Canonicity of pure coeffect-graded CBPV]
  If $\vdash t : \diamond P$ then $t =_{\beta} \ret v$ for some $\vdash v : P$.
\end{conjecture}
This normalization result can be proved with standard techniques such
as reducibility candidates or normalization by evaluation (\cf
\citet{abelSattler:ppdp19}).

@@

>>>>

Coeffect type systems
\citep{brunel:esop14,orchard:icfp14,ghicaSmith:esop14} have been
developed to track resource consumption by attaching usage information
to each variable in the typing context.  Coeffect typing generalizes
linear typing \citep{girard:linear} to quantitative typing
\citep{sergeyVytiniotisPeytonJones:popl14,mcBride:wadler60,atkey:lics18}
and subsumes sensitivity analysis \citep{reedPierce:icfp10} and static
information control flow \citep{volpano:jcs96} aka security typing
\citep{abadiBanerjeeHeintzeRiecke:popl99}.
Coeffects can be modeled by a comonad $\D[r]$ graded over resource
qualifiers $r$ drawn from a preordered semiring.  For the sake of
coeffect-graded CBPV, we generalize graded comonads to graded comonad
coalgebras.  Value types $P$ and contexts $\Gamma$ are then
interpreted as such coalgebras.  Value typing
$\qValTy \gamma \Gamma v P$ and computation typing
$\qCompTy \gamma \Gamma t N$ are based on linear typing an come
equipped with a resource context~$\gamma$.  Semantically, values are
interpreted as morphisms $\Gamma_{r\gamma} \to P_r$ allowing the
``multiplication'' or \emph{scaling} of a value by $r$ % natural in $r$
and computations as morphisms $\Gamma_\gamma \to N$.

@@

\item We further introduce graded comonad coalgebras and a
  coeffect-graded version of CBPV in Section~\ref{sec:coeff}.  We give
  its denotational and operational semantics and adapt Atkey and
  Wood's substitution theorem
  \citeyearpar{atkeyWood:types19,woodAtkey:linearity20} to this
  version of CBPV.

@@

The matrix $\Psi : \R^{\dom \Gamma \times \dom \Delta}$ acts as a
linear map
$(\delta : \R^{\dom \Delta}) \mapsto (\delta\Psi : \R^{\dom \Gamma})$
and allows us to state the substitution theorem.
\begin{theorem}[Substitution preserves typing]
  Let $\SubstTy \Psi \Gamma \sigma \Delta$.
  \begin{enumerate}
  \item If $\qValTy \delta \Delta v P$ then $\pValTy {\delta\Psi} \Gamma {v\sigma} P$.
  \item If $\qCompTy \delta \Delta t N$ then
    $\pCompTy {\delta\Psi} \Gamma {t\sigma} N$.
  \end{enumerate}
\end{theorem}
\begin{proof}
  Straightforward adaptation of \citet{atkeyWood:types19,woodAtkey:linearity20}.
\end{proof}
\begin{corollary}[Single substitution preserves typing]
  \label{cor:tysinglesub}
  \bla
  \begin{enumerate}
  \item If $\qValTy \delta \Gamma v P$ and
      $\ValTy {\qext \gamma \Gamma r x P} w {P'}$
      then $\pValTy {\gamma + r\delta} \Gamma {\qsubst v r x w} {P'}$.
  \item If $\qValTy \delta \Gamma v P$ and
      $\oCompTy {\qext \gamma \Gamma r x P} t {N}$
      then $\pCompTy {\gamma + r\delta} \Gamma {\qsubst v r x t} {N}$.
  % \item If $\qCompTy \delta \Gamma u N$ and
  %     $\oCompTy {\qext \gamma \Gamma r y N} t {N'}$
  %     then $\pCompTy {\gamma + r\delta} \Gamma {\qsubst u r y t} {N'}$.
  \end{enumerate}
\end{corollary}
% %\begin{remark}
% Analogous statements are available for the substitution into values
% $\ValTy {\qext \gamma \Gamma r x P} w {P'}$ and
% $\ValTy {\qext \gamma \Gamma r y N} w {P}$, however, they are not
% invoked in the proof of subject reduction.
% %\end{remark}
\begin{corollary}[Subject reduction]
  $\beta$-reduction of coeffect-graded CBPV preserves types.
\end{corollary}
In the pure (effect-free) version, every computation at value type
returns a value.
\begin{conjecture}[Canonicity of pure coeffect-graded CBPV]
  If $\vdash t : \diamond P$ then $t =_{\beta} \ret v$ for some $\vdash v : P$.
\end{conjecture}
This normalization result can be proved with standard techniques such
as reducibility candidates or normalization by evaluation (\cf
\citet{abelSattler:ppdp19}).

@@

\subsection{Discussion}

Graded CBPV provides syntax to work with graded monads $\T[e]$ and
comonads $\D[r]$.  \emph{A priori}, one could have expected that these
constructs are directly reflected in the syntax of types, as
$\compty e P$ and $\thunkty r N$.  Surprisingly, the adaptation of
graded effect and coeffect typing to CBPV places the qualifiers in the
opposite way, as $\compty r P$ and $\thunkty e N$.  Given our
semantics of types as (co)monad (co)algebras, this apparent oddity has
a natural explanation: the qualifiers do not instantiate the grade in
the monad or comonad, but in the respective (co)algebra.  The
alternative concrete syntax $\diamond P_r$ and $\Box N_e$ would maybe
transport the intended semantics, $\T[-]{\Denp P r}$ and
$\D[-]{\Denn N e}$, more directly.

How does Graded CBPV work as a programming language?  A program is
usually a set of definitions and then an entrypoint in form of an
expression or simply the name of the main procedure.  In Graded CBPV,
the definitions are given as a sequence of let-bindings whose final
body serves as the entrypoint.  The entrypoint should be an expression
of type $\compty 1 P$, producing an observable value and thus, the
effects leading up to this result.  The expressions bound by the lets
will often be functions (or records), but wrapped in thunks
$\thunkty e N$ to satisfy the formal requirement to be of positive
type.  This thunk is where the effects are declared that a function
produces.  Thus, in the big picture, the effects of a computation
\emph{are} stated in the types, not just in the typing judgement
$\qeCompTy \gamma \Gamma t e N$, even though the effect annotation is
in a \emph{a priori} unexpected location.

\section{Related Work}

\subsection{Effect tracking in CBPV}

\citet{mcDermottMycroft:esop19} observe that CBPV can represent
call-by-name and call-by-value evaluation strategies but not
call-by-need.  As a remedy, they extend CBPV to ECBPV by variables
$\underline x : \diamond P$ that hold monadic values, plus a
corresponding let-binding construct.  They further present an
effect-graded version of ECBPV, drawing effect qualifiers from a
preordered monoid.  In contrast to our version, effects annotations
$f$ are stored at monadic types $\compty f P$.  Instead of our
judgement $\CompTy \Gamma t e N$ that accumulates effect traces $e$
and a semantic interpretation of computation types $N$ as graded monad
algebras, they define a syntactic operation $\compty f N$ that pushes
down effect qualifier $f$ to the monadic types $\compty f' P$ in $N$
where the effects are combined to $\compty{f \bu f'} P$.  This way of
handling effect grading seems equivalent to our approach, albeit it is
syntactic rather than semantic.

In previous work, \citet{mcDermottMycroft:ocs18} use coeffect typing
to track variable usage in call-by-need simply-typed lambda calculus
with conditionals.  Compared to the usual resource semiring $\R$
tailored to call-by-name usage analysis, their coeffect algebra is
more expressive: it can produce traces of variable uses, including
bound variables.  They show how graded effect tracking can be
simulated by the coeffect typing, \eg, for the effect of
non-determinism.  Naturally, they focus on operational semantics.

% Filinski’s M3L (MultiMonadic MetaLanguage) [11],
% whose latest version draws heavily on CBPV [12]

To accomodate several sorts of effects in one language (CBPV hosts
only a single effect type), \citet{kammarPlotkin:popl12} present MAIL,
the Multi-Adjunctive Intermediate Language, which is a version of CBPV
\emph{parameterized} by effect qualifiers.  In contrast to our
effect-graded version of CBPV, effect qualifiers are a parameter to
the type system as a whole, basically a \emph{mode} attached to the
typing judgement.  Via effect subsumption, one can switch to a
``wider'' mode.  However, as effects qualifiers are drawn from a
preordered set rather than a monoid, more detailed static effect
traces cannot be captured.  In their own words:
\begin{quotation}
  Instead of one kind of computation, for each effectset
  $\varepsilon \in E$, we have $\varepsilon$-computations
  $\mathsf{Comp}(\varepsilon)$ that can cause effects in
  \ensuremath{\varepsilon}. We view MAIL as multiple copies of CBPV,
  one for each \ensuremath{\varepsilon}, sharing the same values. One
  can translate between these different CBPVs by means of coercion
  [...]
\end{quotation}

% POPL12:
% Algebraic Foundations for Effect-Dependent Optimisations
% Ohad Kammar   Gordon D. Plotkin
% w
% - MAIL: Multi-Adjunctive Intermediate Language
% - Both \Box and \diamonad annotated with effects
% - No effect accumulation

\subsection{Coeffect typing}

Our style of coeffect typing is heavily influenced by
\citet{mcBride:wadler60} and \citet{atkey:lics18} who separate the
quantification of variable \emph{usage} $\gamma$ from the variable
\emph{declaration} in an ordinary typing context $\Gamma$.  This
division of labor removed a major obstacle in the integration of
linear and dependent types: the customary sorting of linear and
unrestricted variables into separate typing contexts, and the
dominance of context concatenation operations in the typing of
multiplicative connectives in linear logic, \eg:
\[
  \ru{\Delta_1 \vdash v_1 : A_1 \qquad
      \Delta_2 \vdash v_2 : A_2
    }{\Delta_1.\Delta_2 \vdash (v_1, v_2) : A_1 \otimes A_2}
\]
In contrast, the coeffect type systems of
\citet{reedPierce:icfp10},
\citet{orchard:icfp14},
\citet{brunel:esop14},
\citet{ghicaSmith:esop14}, and
\citet{orchard:icfp19}
keep context concatenation which does no harm in simply-typed settings
but seems to duplicate a mechanism already present in form of the
resource qualifier $0$.

Our treatment of substitution in coeffect-graded CBPV is based on
\citet{woodAtkey:linearity20}, who formalize (in Agda) substitution
for intuitionistic linear logic with a subexponential representing a
graded comonad.  They observe that instead of a preordered semiring, a
left skew semiring can be used as resource algebra.  In a skew
semiring, the laws involving multiplication only hold as inequalities,
not equalities.

\subsection{Interaction of effects and coeffects}

\citet{orchard:icfp16} investigate systematically the possible
interactions between semiring-graded coeffects and monoid-graded
effects, in form of distributive laws between the graded comonad and
monad.  We do not rely on such distributive laws in Graded CBPV.  The
strict separation of value and computation types and the placement of
the monad and comonad at the transition points makes interaction
optional.  How to integrate distributive laws into Graded CBPV is left
for future research.

\section{Conclusions}

In this article, we have demonstrated that CBPV can accommodate graded
effects and coeffects in a smooth way, keeping the term grammar
virtually unchanged and only adding effect and coeffect annotations at
the transition points between value and computation types.  The
surprising simplicity of our solution speaks for the design quality of
Levy's CBPV calculus \cite{levy:hosc06}.  For instance, the analogy to
intuitionistic linear logic---that lies at the heart of coeffect
typing---is already implicitly present in the separation of value
$(\otimes)$ and computation type products ($\Pi$).

For the semantics of Graded CBPV, we have evolved the concepts of
monad algebra and comonad coalgebra to their graded versions.

In future work, we would like to investigate whether the distributive
laws of \citet{orchard:icfp16} carry over to Graded CBPV, \eg, whether
and how a graded comonad can be distributed over a graded monad
algebra or, in the symmetric case, a graded monad distributes over a
graded comonad coalgebra.
